Carp\'{e} Data: Supporting Serendipitous Data Integration in Personal Information Management

### 2012-09-15 chi2012.tex
revisions to introduction

In recent years, an unprecedented quantity and variety of information has been made available as structured data on the Web, through APIs, datasets, and data feeds.  This information includes access to data previously hidden behind services and applications, such as retailer product catalogues, information not previously directly released to the public, such as open government data or records of personal financial transactions, and new kinds of data generated by emerging kinds of data sources, such as wearable and smartphone-based biosensors.

While a primary goal of opening up such direct access to information is the hope that end-user citizens will make more informed decisions pertaining to their health, wealth, or well-being \cite{}, use of this data beyond specialised app developers, journalists and other Òdata specialistsÓ remains minimal.  Our hypothesis is that the main impediment towards end-user access to this data is a mismatch between the capabilities of tools currently used to satisfy daily information needs, and those required to effectively browse, use and consume heterogeneous structured data from diverse sources.  For example, while much of this data represents a many and varied set of things, activities, and descriptions at various granularities, most commonly used structured personal information management (PIM) tools either can only manage a small, fixed set of data types (such as digital calendaring tools and  to-do list managers) or provide little or no support for structured data at all, such as text editors and sketching/drawing tools \cite{}.

If PIM tools were extended to enable end-users to make effective use of the emerging ecosystem of personal data, how would this impact personal information practice? This paper presents our initial results towards answering this question through a sequential exploratory mixed-method investigation of the personal data integration problem.  We start with a small interview study examining information tasks people perform that rely on multiple sources of information, and how such needs are currently satisfied.  Our results suggest that people prefer to rely upon multiple, diverse sources to singular integrated ones for a number of reasons, including coverage, reliability, and seeking out more suitable alternatives.  Second, we conduct a analysis of various popular personal data sources available today to identify barriers to effective unification of data from each.  We draw upon definitions and work from the database integration systems literature to characterise types of heterogeneity exhibited across data sources in three popular domains: social networking, shopping, and dining, finding that terminological heterogeneity dominates the simpler data feeds (including social networking and restaurant recommendations), while structural issues pervade the more complex schemas of online retailers' product catalogues.

Based upon the data integration needs revealed by people in the interviews, and the kinds of heterogeneity observed in the data feeds, we embarked on a user-centric design exercise towards an interface to facilitate simple, ``light-touch'' integration tasks of diverse, heterogeneous data sources.  We describe our initial foray into this exercise, an interface we call DataPalette, which focuses on the most common types of heterogeneity to enable serendipitous ``data mixing'' - a form of simple integration sufficient to let people easily and effectively combine and compare information sources without the need to write code.  Our usability evaluation of DataPalette reveals that most users were comfortable with and that this interface is effectively expedites the most sophisticated of the typical kinds of ``data integration'' tasks people need to perform while planning personal,

@@

\section{Designing DataPalette : An Interface for Serendipitous Data Mixing}
\begin{figure*}[htbp]
\begin{center}
\includegraphics[width=18cm]{img/screenshot}
\caption{DataPalette Workspace, a file manager-inspired Drag and Drop interface for structured data mixing. Collapsable displays of data sources are listed on the left, from which instances or entire sets of items can be dragged to form a windoow (group) on the workspace.}
\label{fig:workspace}
\end{center}
\end{figure*}

At a high level, the overall goal of DataPalette was to empower end-users to mix arbitrary structured data feeds on-the-fly while solving their particular information task(s) - be them sensemaking or decision-making related - in order to more effectively accomplish these tasks.  Unlike Mashup makers, including Yahoo Pipes!, the goal was to facilitate such integration without a need for an explicit separate mapping or integration step, which would detract from the overall usability or usefulness of the system by making people do something that did not immediately contribute to their resolving their need.

The two prestudies provided valuable insight towards this goal that allowed us to guide our design process. The first prestudy confirmed the need for data integration by showing that people did, in fact, prefer to rely upon multiple sources of information for making important decisions, and identified the kinds of sources people most commonly used.  The second prestudy revealed that the most common inconisistencies that occurred in the APIs and data feeds available through from such data sources consisted predominantly of the simpler kinds of heterogeneity - terminological and simpler forms of structural heterogeneity, rather than the delicate semantic/modeling variety, which were comparatively rare. This gave us optimism that an interactional approach optimised around the these simpler kinds of heterogeneity would be effective.

In the following sections we describe the design process we used to create prototype DataPalette, followed by a detailed description of the design of the prototype and justification.

\subsection{Process}
We followed a five-phase spiral model consisting of planning, designing, mocking-up, prototyping, and testing to derive the design of DataPalette.  This process was used because of the high degree of initial uncertainty associated with the mechanisms for facilitating the various aspects of effective data mixing that we describe in the next section. We approached this design challenge through alternative generation and elimination; alternatives were designed, drawn up and evaluated by expert colleagues (and in later iterations, non-expert potential users), who provided critiques of the alternatives. The ease with which designs could be prototyped in HTML5 with modern web frameworks allowed us to perform this develop-test-iterate rapidly through seven iterations of this process.

\subsection{Interaction mechanisms for data mixing}
In order to familiar to new users, we borrowed the overall interaction and feel from MacOS Finder-style file managers, with support for separate windows which can be freely moved, resized and titled. (We later discovered that windows caused considerable clutter at the later stages of some tasks, a fact we did not detect until the main study as we describe later.)  Instead of files, however, the element of manipulation was a single, discrete structured data item, or \emph{instance}.  In order to define \emph{instances} for non-RDF data, such as JSON or XML, we defined all subtrees with more than 1 child as its an instance.  Data sources were represented as collapsable panels in a sidebar on the left of the workspace, from which instances or entire data sources could be ``pulled off'' onto the workspace.  Figure \

As we refined our designs, we arrived at four simple mechanisms to support data mixing.  These were combined with a number of simple interaction mechanisms that were borrowed from other systems

The ultimatel design of datapal

@@

>>>>

In recent years, an unprecedented quantity and variety of information has been made available as structured data on the Web, through APIs, datasets, and data feeds.  This information includes: access to data previously hidden behind services and applications, such as retailer product catalogues; information not previously directly released to the public, such as open government data or records of personal financial transactions; and new kinds of data generated by emerging kinds of data sources, such as wearable and smartphone-based biosensors.  While a primary goal of opening up such direct access to information is that end-user citizens will make more informed decisions pertaining to their personal information, for example health, wealth, and well-being \cite{}.  This data is predominately used by  specialised app developers, journalists and other ``data specialists'' but remains inaccessible to citizens because of the available tools.  For example, the most commonly used structured personal information management (PIM) tools either manage a small, fixed set of data types (such as digital calendaring tools and  to-do list managers) or provide little or no support for structured data at all, such as text editors and sketching/drawing tools \cite{}.

We hypothesise that end-users would benefit from this rich data with tools allowing them to effectively browse, use and consume heterogeneous structured data from diverse sources.  In this paper, we aim to investigate how extending PIM tools to effectively use emerging ecosystems of personal data would impact information practice.  We use a three stage sequential exploratory mixed method approach: First, we performed a qualitative analysis, using a small semi-structured interview study to examine the types of tasks people perform requiring more than one source of information, and the processes that they would use to perform such a task.  Our results suggest that people prefer to rely upon multiple, diverse sources to singular integrated ones for a number of reasons, including coverage, reliability, and seeking out more suitable alternatives.

Second, we conducted a quantitative analysis of various popular personal data sources available today to identify potential barriers to effective unification of data from each.  We then draw upon definitions and work from the database integration systems literature to characterise types of heterogeneity exhibited across data sources in three popular domains: social networking, shopping, and dining, finding that terminological heterogeneity dominates the simpler data feeds (including social networking and restaurant recommendations), while structural issues pervade the more complex schemas of online retailers' product catalogues.

Third, we developed DataPalette, which focuses on the most common types of heterogeneity to enable serendipitous ``data mixing'' - a form of simple integration sufficient to allow people to easily and effectively combine and compare information sources without the need to write code.  It was principally designed based upon people's data integration needs and to support data heterogeneity observed in the previous studies.  We embarked on a user-centric design exercise towards an interface to facilitate simple, ``light-touch'' integration tasks of diverse, heterogeneous data sources.  We then performed a usability evaluation of DataPalette, which revealed that most users were comfortable with and that this interface is effectively expedites the most sophisticated of the typical kinds of ``data integration'' tasks people need to perform while planning personal.

@@

\section{Supporting Design of the DataPalette}

\subsection{Design Goals}

At a high level, the overall goal of DataPalette was to empower end-user citizens to mix arbitrary structured data feeds on-the-fly while in the process of doing their particular information task - be it sensemaking or decision-making related.  Unlike Mashup makers, including Yahoo Pipes!, our hope was that this could be accomplished without need for a separate, explicit mapping, integration or manual integration step which would make the effort-overhead and its use so high as to impede its use as a PIM tool.

Thus we wanted to focus on the most common

The second prestudy highlighted the most common inconisistencies that occurred among data sources and allowed us to the focus on a smaller set of problems. The most common examples were terminological (labeling) inconsitencies, minor structural inconsistencies (pertaining to values at different levels of a field's record), and value range terminological An additional problem that needed to be addressed for effectively mixing data from multipl

\emph{coreference}

What are the problems observed needed to address
	What kind of interactions can be designed to solve them?
\subsection{Process}

We followed a spiral model consisting of the phases of plan, design, mock-up, prototype, test, re-plan, and re-design to derive the design of DataPalette.  This process was used because of the high degree of initial uncertainty associated with how to design easy to use

@@

\begin{figure*}[htbp]
\begin{center}
\includegraphics[width=18cm]{img/dragdrop}
\caption{WebBox uses a Drag and Drop interface to add data into the workspace.}
\label{fig:dragdrop}
\end{center}
\end{figure*}
